Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-02-24 paper/sec/abstract.tex
Paper: Minor edits.   * Use central place for LOCs.

\begin{abstract}
Estimating the number of distinct elements in a data stream is a classical problem with numerous applications in computer science.
We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (called the CVM algorithm).
Their algorithm deviated considerably from the state of the art, due to its avoidance of intricate derandomization techniques, while still maintaining a close-to-optimal logarithmic space complexity.

Central to our formalization is a new proof technique based on \emph{functional probabilistic invariants}, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method.
This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.
Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
In this paper, we introduce the proof technique, describe its use in mechanizing both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.
\end{abstract}

>>>>

\begin{abstract}
Estimating the number of distinct elements in a data stream is a classic problem with numerous applications in computer science.
We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (called the CVM algorithm).
Their algorithm deviated considerably from the state of the art, due to its avoidance of intricate derandomization techniques, while still maintaining a close-to-optimal logarithmic space complexity.

Central to our formalization is a new proof technique based on \emph{functional probabilistic invariants}, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method without relying on independence.
This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.
Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
In this paper, we introduce the proof technique, describe its use in mechanizing both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.
\end{abstract}

### 2025-02-24 paper/sec/conclusion.tex
Paper: Minor edits.   * Use central place for LOCs.

\section{Conclusion}\label{sec:conclusion}
We presented the first formalization of the CVM algorithm using Isabelle/HOL.
Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method.
Comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our new invariant-based technique yields a considerably shorter formalization (with 1004 vs.~2646 lines).
Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.

>>>>

\section{Conclusion}\label{sec:conclusion}
We presented the first formalization of the CVM algorithm using Isabelle/HOL.
Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method.
Comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our new invariant-based technique yields a considerably shorter formalization (with \locnew vs. \locold~lines).
Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.

### 2025-02-24 paper/sec/intro.tex
Paper: Minor edits.   * Use central place for LOCs.

We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
As a rough point of comparison, verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle/HOL~\cite{nipkow2002}, while the original proof required \todo{x} lines.

@@

The supplementary material contains:
\begin{itemize}
\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants};
\item formalization of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
\item formalization of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
\end{itemize}

>>>>

We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
As a rough point of comparison, verification of the CVM algorithm using our new technique requires only \locnew~lines in Isabelle/HOL~\cite{nipkow2002}, while the original proof required \locold~lines.

@@

The supplementary material contains:
\begin{itemize}
\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using probabilistic functional invariants;
\item formalization of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
\item formalization of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
\end{itemize}
